### 2016-06-26 ansi-isolation.tex
Figures added to 3.2. Edits done.

(\A,\visZ) \Vdash \eta \in T_i & \Leftrightarrow & \eta \in \A \conj \txn(\eta) = T_i\\
(\A,\visZ) \Vdash \eta_1 \visar \eta_2 & \Leftrightarrow & \{\eta_1,\eta_2\}

@@

(\A,\visZ) \Vdash \eta_1 \soar \eta_2 & \Leftrightarrow & \{\eta_1,\eta_2\}

@@

\E \Vdash S \subseteq T_i & \Leftrightarrow & \forall \eta.~ \eta

@@

\E \Vdash T_i \visar \eta & \Leftrightarrow &\forall\eta_1

@@

\underE{T_i \visar T_j} & \Leftrightarrow &  \forall\eta_1,\eta_2.\,

@@

\underE{T_i \invisar T_j} & \Leftrightarrow &  \forall\eta_1,\eta_2.\,

@@

\underE{T_i \wrstoar X} & \Leftrightarrow & \exists\eta.~

@@

\underE{\C{RMWVis}(T_j)} & \Leftrightarrow & \forall\eta_1,\eta_2.\,

@@

\underE{\C{MonotonicVis}(T_j)} & \Leftrightarrow & 

@@

%  \C{CausalVis}(T_i) & \Leftrightarrow & 

@@

\underE{\C{AtomicVis}(T_j)} & \Leftrightarrow & 

@@

\underE{\C{CommitVis}(T_j)} & \Leftrightarrow & 

@@

% \underE{\C{TransVis}(T_j)} & \Leftrightarrow &  \forall

@@

\underE{\C{SnapshotVis}(T_i,T_j)} & \Leftrightarrow &  \underE{T_i

@@

\underE{\C{OneWaySER}(T_i,T_j)} & \Leftrightarrow &  \underE{T_i
       \C{COMMIT} \Rightarrow \underE{T_j \visar \eta})\\
\underE{\C{RC}(T_j)} & \Leftrightarrow & \underE{\C{AtomicVis}(T_j)} 
\underE{\C{MAV}(T_j)} & \Leftrightarrow & \underE{\C{RC}(T_j)} \conj
\underE{\C{RR}(T_j)} & \Leftrightarrow & \underE{\C{MAV}(T_j)}

@@

\underE{\C{SI}(T_j)} & \Leftrightarrow &  \underE{\C{RR}(T_j)}

@@

  &   & \hspace*{2in} \Rightarrow \underE{\C{OneWaySER}(T_i,T_j)}\\
\underE{\C{SER}(T_j)} & \Leftrightarrow & \underE{\C{MAV}(T_j)}
       \conj \forall T_i.\,T_i \neq T_j 
       \Rightarrow \underE{\C{OneWaySER}(T_i,T_j)}\\
% &   & \hspace*{2in}\conj  \C{TotalVis}(T_i,T_j)\\

@@

In the context of a trace $(\A,\visZ)$, an effect $\eta$ is said to
belong to a transaction $T_i$ if $\eta$ belongs to the effect set $A$
and its transaction identifier is $T_i$. The containment relation is
trivially lifted to the set of effects to define $\underE{S \subseteq
T_i}$.  The interpretation of $\eta_1 \visar \eta_2$ and $\eta_1
\soar \eta_2$ are straightforward and explained in \S\ref{sec:syntax}.
A transaction $T_i$ is said to be visible to an effect $\eta$ if every
effect $\eta_1$ of $T_i$ recorded by the trace is visible to $\eta$.
$T_i$ may be visible to $\eta$ but may not be visible to every other
effect in the transaction. For a transaction $T_i$ to be considered to
be visible to a transaction $T_j$ in the context of a trace $\E$
(written $\underE{T_i \visar T_j}$), every effect ($\eta_1$) of $T_i$
present in $\E$ must be visible to every effect ($\eta_2$) of $T_j$ in
$\E$. Conversely, if none of the effects of $T_i$ present in $\E$ are
visible to any effect of $T_j$, then $T_i$ is considered invisible to
$T_j$ under $\E$ (written $\underE{T_i \invisar T_j}$). Transaction
$T_i$ is said to have written to a variable $X$ under $\E$ (i.e.,
$\underE{T_i \wrstoar X}$) if there exists a write-to-$X$ effect from
$T_i$ in $\E$.

Various isolation guarantees are defined as propositions indexed by
a transaction identifier $T_j$. Transaction $T_j$ is said to have
experienced \emph{read-my-writes} visibility under $\E$ if every
effect ($\eta_1$) of $T_j$ is visible to every subsequent effect
($\eta_2$) of the same transaction in $\E$. This lets $T_j$ to never
lose its own updates. Monotonic visibility adds one more constraint to
read-my-writes; besides requiring $\eta_1$ to be visible to $\eta_2$,
it also requires every effect ($\eta$) visible to $\eta_1$ in $\E$ to
be visible to $\eta_2$ as well. Thus $T_j$ witnesses monotonically
increasing state as time progresses. Atomic visibility allows an
effect $\eta_2$ of $T_j$ to witness an effect $\eta_1$ of $T_i$ only
if all effects of $T_i$ in $\E$ are also visible to $\eta_2$. Atomic
visibility thus prevents a transaction from being partially visible.
However, atomic visiblity does not prevent an uncommitted transaction
from being visible. To confine visibility to only committed
transactions, we need an additional constraint that requires an effect
($\eta_2$) of $T_j$ to also witness the commit effect of $T_i$
whenever it witnesses some effect ($\eta_1$) of $T_i$. This additional
constraint is captured by the $\mathtt{CommitVis}$ specification. A
transaction $T_i$ is said to be snapshot-visible to a transaction
$T_j$ if either it is visible to $T_j$, or it is invisible; it is
forbidden for only a suffix (more generally, a subset) of $T_j$ to
witness $T_i$. Observe that snapshot visibility permits $T_i$ and $T_j$ to
execute and commit while being  oblivious of each other. One-way
serializability proscribes this possiblity. $T_i$
is said to be one-way-serializable w.r.t $T_j$ if either it is visible
to $T_j$ or $T_j$ is visible to the commit effect of $T_i$.  Note that
both $\mathtt{SnapshotVis}$ and $\mathtt{OneWaySER}$ are asymmetric
definitions that guarantee complete visiblity or invisibility of $T_i$
to $T_j$, but not the converse. While $\mathtt{SnapshotVis}$ provides
no guarantees to $T_i$, $\mathtt{OneWaySER}$ guarantees that at least
the commit effect of $T_i$ witnesses $T_j$.

The ANSI SQL 92 standard requires \iso{Read Committted} isolation to
avoid the dirty reads phenomenon, which is achieved by enforcing
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
specification is therefore a combination of these two guarantees. The
specification also agrees with the description and
implementation~\cite{bailishat,pldi15} of {\sc rc} for highly
available replicated stores. On relational databases, however, {\sc
rc} has also come to be associated with the $\mathtt{MonotonicVis}$
guarantee.  Nonetheless, $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$
are sufficient to reason about {\sc rc} isolation on relational stores
too. The combination of these guarantees with the {\sc sc} property of
relational stores (formalized in \S\ref{sec:store-consistency})
automatically leads to the monotonicity guarantee, which probably
explains why {\sc rc} comes with $\mathtt{MonotonicVis}$ on such
stores. On weakly consistent stores however, $\mathtt{AtomicVis}$ and
$\mathtt{CommitVis}$ do not imply $\mathtt{MonotonicVis}$. A stronger
isolation level called \iso{Monotonic Atomic View} ({\sc mav} of
Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} was proposed to
explicitly extend {\sc rc} with monotonicity on such stores. Sample
{\sc rc} executions shown in Fig.~\ref{fig:rc-executions} are also
{\sc mav}. \iso{Repeatable Read} spec ({\sc rr}) extends {\sc mav}
with the snapshot visibility guarantee as described before.
\iso{Snapshot Isolation} spec ({\sc si}) extends {\sc rr} with a
one-way serializability guarantee w.r.t the transactions that perform
conflicting writes (i.e., writes to the same shared variable).
Finally, \iso{Serializable} isolation extends one-way
$\mathtt{SnapshotVis}$ to all transactions, including those that do
not conflict\footnote{Fig.~\ref{fig:ansi-isolation} presents slightly
weaker versions of {\sc si} and {\sc ser} specs in the interest of
clarity.}. \GK{ToDo: Figures. Need to figure out a convention to
uniformly represent various isolation levels. Current idea: concurrent
timelines with arrows in both directions.}

>>>>

(\A,\visZ) \Vdash \eta \in T_i & \defeq & \eta \in \A \conj \txn(\eta) = T_i\\
(\A,\visZ) \Vdash \eta_1 \visar \eta_2 & \defeq & \{\eta_1,\eta_2\}

@@

(\A,\visZ) \Vdash \eta_1 \soar \eta_2 & \defeq & \{\eta_1,\eta_2\}

@@

\E \Vdash S \subseteq T_i & \defeq & \forall \eta.~ \eta

@@

\E \Vdash T_i \visar \eta & \defeq &\forall\eta_1

@@

\underE{T_i \visar T_j} & \defeq &  \forall\eta_1,\eta_2.\,

@@

\underE{T_i \invisar T_j} & \defeq &  \forall\eta_1,\eta_2.\,

@@

\underE{T_i \wrstoar X} & \defeq & \exists\eta.~

@@

\underE{T_i \rdsfmar X} & \defeq & \exists\eta.~
        \underE{\eta \in T_i} \conj \kind(\eta) = \C{RD}(X)\\
\underE{T_i \usesar X} & \defeq & \underE{T_i \wrstoar X} \disj
      \underE{T_i \rdsfmar X}\\
\underE{\C{RMWVis}(T_j)} & \defeq & \forall\eta_1,\eta_2.\,

@@

\underE{\C{MonotonicVis}(T_j)} & \defeq & 

@@

%  \C{CausalVis}(T_i) & \defeq & 

@@

\underE{\C{AtomicVis}(T_j)} & \defeq & 

@@

\underE{\C{CommitVis}(T_j)} & \defeq & 

@@

% \underE{\C{TransVis}(T_j)} & \defeq &  \forall

@@

\underE{\C{RC}(T_j)} & \defeq & \underE{\C{AtomicVis}(T_j)} 
\underE{\C{MAV}(T_j)} & \defeq & \underE{\C{RC}(T_j)} \conj
\underE{\C{SnapshotVis}(T_i,T_j)} & \defeq &  \underE{T_i

@@

\underE{\C{RR}(T_j)} & \defeq & \underE{\C{MAV}(T_j)}

@@

\underE{\C{SnapshotSER}(T_i,T_j,X)} & \defeq &  \underE{T_i
       \C{WR(X)} \Rightarrow \underE{T_j \visar \eta})\\
\underE{\C{SI}(T_j)} & \defeq &  \underE{\C{RR}(T_j)}

@@

  &   & \hspace*{2in} \Rightarrow \underE{\C{SnapshotSER}(T_i,T_j,X)}\\
\underE{\C{SER}(T_j)} & \defeq & \underE{\C{RR}(T_j)}
        \conj \forall T_i.\,(T_i \neq T_j \conj \exists X.\, 
        \underE{T_i \wrstoar X} \conj \underE{T_j \usesar X})\\
  &   & \hspace*{2in} \Rightarrow \underE{\C{SnapshotSER}(T_i,T_j,X)}\\

@@

In the context of a trace $(\A,\visZ)$, an effect $\eta$ is said to
belong to a transaction $T_i$ if $\eta$ belongs to the effect set $A$
and its transaction identifier is $T_i$. The containment relation is
trivially lifted to the set of effects to define $\underE{S \subseteq
T_i}$.  The interpretation of $\eta_1 \visar \eta_2$ and $\eta_1
\soar \eta_2$ are straightforward and explained in \S\ref{sec:syntax}.
A transaction $T_i$ is said to be visible to an effect $\eta$ if every
effect $\eta_1$ of $T_i$ recorded by the trace is visible to $\eta$.
$T_i$ may be visible to $\eta$ but may not be visible to every other
effect in the transaction. For a transaction $T_i$ to be considered to
be visible to a transaction $T_j$ in the context of a trace $\E$
(written $\underE{T_i \visar T_j}$), every effect ($\eta_1$) of $T_i$
present in $\E$ must be visible to every effect ($\eta_2$) of $T_j$ in
$\E$. Conversely, if none of the effects of $T_i$ present in $\E$ are
visible to any effect of $T_j$, then $T_i$ is considered invisible to
$T_j$ under $\E$ (written $\underE{T_i \invisar T_j}$). Transaction
$T_i$ is said to have written to a variable $X$ under $\E$ (i.e.,
$\underE{T_i \wrstoar X}$) if there exists a write-to-$X$ effect from
$T_i$ in $\E$. \emph{Reads-from} is similar. $T_i$ \emph{uses} $X$ if
it reads or writes $X$.

Various isolation guarantees are defined as propositions indexed by
a transaction identifier $T_j$. Transaction $T_j$ is said to have
experienced \emph{read-my-writes} visibility under $\E$ if every
effect ($\eta_1$) of $T_j$ is visible to every subsequent effect
($\eta_2$) of the same transaction in $\E$. This lets $T_j$ to never
lose its own updates. Monotonic visibility adds one more constraint to
read-my-writes; besides requiring $\eta_1$ to be visible to $\eta_2$,
it also requires every effect ($\eta$) visible to $\eta_1$ in $\E$ to
be visible to $\eta_2$ as well. Thus $T_j$ witnesses monotonically
increasing state as time progresses. Atomic visibility allows an
effect $\eta_2$ of $T_j$ to witness an effect $\eta_1$ of $T_i$ only
if all effects of $T_i$ in $\E$ are also visible to $\eta_2$. Atomic
visibility thus prevents a transaction from being partially visible.
However, atomic visiblity does not prevent an uncommitted
transaction from being visible. This is addressed by \C{CommitVis},
which requires the commit effect of $T_i$ to be visible whenever any
effect of $T_i$ is visible.

ANSI SQL 92 standard requires \iso{Read Committted} isolation to avoid
the dirty reads phenomenon, which is achieved by enforcing
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. {\sc rc}
specification is therefore a combination of these two guarantees. The
specification also agrees with the description and
implementation~\cite{bailishat,pldi15} of {\sc rc} for highly
available replicated stores. On relational databases, however, {\sc
rc} has also come to be associated with the $\mathtt{MonotonicVis}$
guarantee.  Nonetheless, $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$
are sufficient to reason about {\sc rc} isolation on relational stores
too. The combination of these guarantees with the {\sc sc} property of
relational stores (formalized in \S\ref{sec:store-consistency})
automatically leads to the monotonicity guarantee, which explains why
{\sc rc} comes with $\mathtt{MonotonicVis}$ on such stores regardless
of the implementation. On weakly consistent stores however,
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ do not imply
$\mathtt{MonotonicVis}$. A stronger isolation level called
\iso{Monotonic Atomic View} ({\sc mav} of
Fig.~\ref{fig:ansi-isolation})~\cite{bailishat,pldi15} was proposed to
explicitly extend {\sc rc} with monotonicity on such stores. Sample
{\sc rc} executions shown in Fig.~\ref{fig:rc-executions} are also
{\sc mav}.

\begin{figure}
\centering
\subcaptionbox {
  {\sc rr}($T_1$), {\sc mav}($T_2$).
  \label{fig:ansi-iso-eg-rr}
} [
  0.33\columnwidth
] {
  \includegraphics[scale=0.36]{Figures/rr-eg}
}
\subcaptionbox {
  {\sc si}($T_1$), {\sc mav}($T_2$)
  \label{fig:ansi-iso-eg-si}
} [
  0.33\columnwidth
] {
  \includegraphics[scale=0.36]{Figures/si-eg-2}
}
\subcaptionbox {
  {\sc ser}($T_3$), {\sc mav}($T_4$)
  \label{fig:ansi-iso-eg-ser}
}{
  \includegraphics[scale=0.32]{Figures/ser-eg}
}

\caption{Sample executions of concurrent transactions at different
isolation levels. Solid (black) arrows indicate the timeline. Points
on the timeline mark the time when an operation is executed. Dotted
(blue) arrows denote $\visZ$. }
\label{fig:ansi-iso-eg}
\end{figure}

A transaction $T_i$ is said to be snapshot-visible to a transaction
$T_j$ if either it is visible to $T_j$, or it is invisible; it is
forbidden for only a suffix (more generally, a subset) of $T_j$ to
witness $T_i$. The specification of\iso{Repeatable Read} ({\sc rr})
isolation extends {\sc mav} with the snapshot visibility guarantee.
Observe that snapshot visibility permits $T_i$ and $T_j$ to execute
and commit while being oblivious of each other. This scenario is
captured in Fig.~\ref{fig:ansi-iso-eg-rr}, where transactions $T_1$
and $T_2$, which perform conflicting writes, execute against a
snapshot of the database and commit concurrently.

\iso{Snapshot Isolation} ({\sc SI}) proscribes this possibility. If
$T_i$ and $T_j$ both write to the same shared variable, then {\sc si}
insists that either $T_i$ be visible to $T_j$, or $T_j$ be visible to
the conflicting write of $T_i$ (this is captured by the auxiliary
definition \C{SnapshotSER}). Fig.~\ref{fig:ansi-iso-eg-si} shows an
execution where {\sc si} transaction $T_1$ witnesses a snapshot of the
database that doesn't include $T_2$, but the conflicting write to $X$
in $T_2$, and the subsequent operations (due to the monotonicity
requirement of {\sc mav}) witness $T_1$. Thus, $T_2$ is aware of $T_1$
when it committed. Lastly, \iso{Serializable} isolation extends
$\mathtt{SnapshotSER}$ to also cover non-conflicting transactions that
write to variables read in the current transaction.
Fig.~\ref{fig:ansi-iso-eg-ser} shows an execution of an {\sc ser}
transacton $T_3$ and a {\sc mav transaction} $T_4$, which write to $X$
and $Y$, respectively. The snapshot witnessed by the {\sc ser}
transaction $T_3$ does not include $T_4$, but the write to $Y$ in
$T_4$, although non-conflicting, witnesses $T_1$ because $Y$ is read
by $T_1$. Visibility includes other operations of $T_4$ because of
{\sc mav}.

% Note that both $\mathtt{SnapshotVis}$ and $\mathtt{OneWaySER}$
% are asymmetric definitions that guarantee complete visiblity or
% invisibility of $T_i$ to $T_j$, but not the converse. While
% $\mathtt{SnapshotVis}$ provides no guarantees to $T_i$,
% $\mathtt{OneWaySER}$ guarantees that at least the commit effect of
% $T_i$ witnesses $T_j$.

% \iso{Snapshot Isolation} spec ({\sc si}) extends {\sc rr} with a
% one-way serializability guarantee w.r.t the transactions that perform
% conflicting writes (i.e., writes to the same shared variable).
% \footnote{Fig.~\ref{fig:ansi-isolation} presents slightly
% weaker versions of {\sc si} and {\sc ser} specs in the interest of
% clarity.}. Fig.~\ref{fig:ansi-iso-eg} shows sample executions of
% transactions $T_1$ and $T_2$. Both transactions read and write to
% shared variables $X$ and $Y$. In the first execution, $T_2$ commits
% while $T_1$ is still in progress, but {\sc rr} isolation prevents
% $T_1$ from witnessing the writes of $T_2$. In the second execution,
